---
title: "Exercises using packages bigsnpr and bigstatsr"
author: "Florian Privé"
date: "January 22, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

In this document, I show how to use some of the features of packages {bigsnpr} and {bigstatsr}.
Note that many functions used here come from package {bigstatsr} and could therefore be used on other data encoded as matrix-like (outside of the field of genotype data).

## Get data

```{r setup, include=FALSE}
unlink(paste0("tmp-data/public-data", c(".bk", ".rds")))
options(htmltools.dir.version = FALSE, width = 65)
knitr::opts_knit$set(global.par = TRUE)
knitr::opts_chunk$set(echo = TRUE, fig.align='center', dev='png', dpi = 95)
```

Download [data](https://github.com/privefl/bigsnpr/raw/master/data-raw/public-data.zip) and unzip files. I store those files in a directory called `"tmp-data"` here.

You can see [there](https://github.com/privefl/bigsnpr/blob/master/data-raw/public-data.R) how I generated these data from [the 1000 Genomes project](https://www.nature.com/articles/nature15393).

## Read from the PLINK files

```{r}
# Load packages bigsnpr and bigstatsr
library(bigsnpr)
# Read from bed/bim/fam, it will create new files.
snp_readBed("tmp-data/public-data.bed")
# Attach the "bigSNP" object in R session
obj.bigSNP <- snp_attach("tmp-data/public-data.rds")
# See how it looks like
str(obj.bigSNP, max.level = 2, strict.width = "cut")
# Get aliases for useful slots
G   <- obj.bigSNP$genotypes
CHR <- obj.bigSNP$map$chromosome
POS <- obj.bigSNP$map$physical.pos
y   <- obj.bigSNP$fam$affection
sex <- obj.bigSNP$fam$sex
pop <- obj.bigSNP$fam$family.ID
NCORES <- nb_cores()
# Check some counts for the 10 first SNPs
big_counts(G, ind.col = 1:10)
```

## What you must do

You need to 

- Explore the data

- Assess population structure of the data

- Find the variables associated with the disease status (`y`)

- Predict the disease status using the following indices as training/test sets:

    ```{r}
    # Divide the indices in training/test sets
    set.seed(1)
    ind.train <- sample(nrow(G), 400)
    ind.test <- setdiff(rows_along(G), ind.train)
    ```

For this, you can use whatever tools you want because the data is quite small. 
In the following section, I give some (scalable) solutions using using packages {bigstatsr} and {bigsnpr}.

## Solution using {bigstatsr} and {bigsnpr}

### Population structure: Principal Component Analysis

Let us compute first principal components of the scaled genotype matrix:

```{r}
# Compute partial SVD (10 PCs by default) using random projections
# big_scale() computes means and standard deviations for scaling
svd <- big_randomSVD(G, big_scale(), ncores = nb_cores())
```

```{r}
# Scree plot
plot(svd)
library(ggplot2)
# Scores plot + color for population
plot(svd, type = "scores") +
  aes(color = pop)
plot(svd, type = "scores", scores = 3:4) +
  aes(color = pop)
```

```{r, fig.asp=1.4}
# Loadings (effects of each variable for each PC)
plot(svd, type = "loadings", loadings = 1:10, coeff = 0.4)
```

Looking at the loadings, we can see that the PCA captures some variation due to large correlation between variables. To learn more about this possible pitfall, please look at [this vignette](https://privefl.github.io/bigsnpr/articles/how-to-PCA.html).

### Association: Genome-Wide Association Study (GWAS)

```{r}
# Association of each variable of `G` with `y` (adjusting for 10 PCs)
gwas <- big_univLogReg(G, y, covar.train = svd$u, ncores = nb_cores())
# Histogram of p-values
plot(gwas)
# Q-Q plot
plot(gwas, type = "Q-Q") + xlim(1, NA)  # snp_qq(gwas) + xlim(1, NA)
# Manhattan plot
snp_manhattan(gwas, CHR, POS, npoints = 20e3) +
  geom_hline(yintercept = -log10(5e-8), color = "red")
```

### Polygenic Risk Score (PRS) with Penalized Logistic Regression (PLR)

$$\arg\!\min_{\beta_0,~\beta}(\lambda, \alpha)\left\{ \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function} + \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} \right\}$$

where

- $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$

- $x$ is denoting the genotypes and covariables (e.g. principal components), 

- $y$ is the disease status we want to predict, 

- $\lambda$ is a regularization parameter that needs to be determined and

- $\alpha$ determines relative parts of the regularization $0 \le \alpha \le 1$. 

<br>

If you want to learn more about our implementation of PLR, please look at [this paper](https://doi.org/10.1101/403337).

<br>

```{r}
# Penalized logistic regression for many alphas and lambdas
mod <- big_spLogReg(G, y[ind.train], ind.train, covar.train = svd$u[ind.train, ],
                    K = 5, alphas = 10^(-(0:4)), ncores = nb_cores())
```

```{r}
# Plot regularization paths (from high lambda to low lambda) 
# for each validation set (color) and each alpha (facet)
plot(mod)
# Get summaries of models
summary(mod)
```

```{r}
# Get the predictions for the test set
pred <- predict(mod, G, ind.test, covar.row = svd$u[ind.test, ])
# Assess the Area Under the ROC Curve
AUC(pred, y[ind.test])
```
