---
title: "Computing polygenic scores using LDpred2"
author: "Florian Privé"
date: "April 2, 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

In this document, we show how to compute polygenic risk scores using LDpred2.

## Downloading genotype data and summary statistics

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, width = 75)
knitr::opts_knit$set(global.par = TRUE, root.dir = "..")
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dev = 'png')
```

You can download [data](https://github.com/privefl/bigsnpr/raw/master/data-raw/public-data.zip) and unzip files in R. We store those files in a directory called `"tmp-data"` here.

```{r, echo=FALSE, eval=FALSE}
unzip("data-raw/public-data.zip")
```

```{r, echo=FALSE}
unlink(paste0("tmp-data/public-data", c(".bk", ".rds")))
```

You can see [there](https://github.com/privefl/bigsnpr/blob/master/data-raw/public-data.R) how we generated these data from [the 1000 Genomes project](https://www.nature.com/articles/nature15393).

First, you need to read genotype data from the PLINK files (or BGEN files) as well as the text file containing summary statistics.

```{r}
# Load packages bigsnpr and bigstatsr
library(bigsnpr)
# Read from bed/bim/fam, it generates .bk and .rds files.
snp_readBed("tmp-data/public-data.bed")
# Attach the "bigSNP" object in R session
obj.bigSNP <- snp_attach("tmp-data/public-data.rds")
# See how the file looks like
str(obj.bigSNP, max.level = 2, strict.width = "cut")
# Get aliases for useful slots
G   <- obj.bigSNP$genotypes
CHR <- obj.bigSNP$map$chromosome
POS <- obj.bigSNP$map$physical.pos
y   <- obj.bigSNP$fam$affection - 1
NCORES <- nb_cores()
# Read external summary statistics
sumstats <- bigreadr::fread2("tmp-data/public-data-sumstats.txt")
str(sumstats)
```

We split genotype data using part of the data to learn parameters of stacking and another part of the data to evaluate statistical properties of polygenic risk score such as AUC. Here we consider that there are 400 individuals in the training set.

```{r}
set.seed(1)
ind.val <- sample(nrow(G), 400)
ind.test <- setdiff(rows_along(G), ind.val)
```

## Matching variants between genotype data and summary statistics 

To match variants contained in genotype data and summary statistics, the variables `"chr"` (chromosome number), `"pos"` (genetic position), `"a0"` (reference allele) and `"a1"` (derived allele) should be available in the summary statistics and in the genotype data. These 4 variables are used to match variants between the two data frames.

```{r}
sumstats$n_eff <- 4 / (1 / sumstats$n_case + 1 / sumstats$n_control)
sumstats$n_case <- sumstats$n_control <- NULL
names(sumstats) <- c("chr", "rsid", "pos", "a0", "a1", "beta", "beta_se", "p", "n_eff")
map <- obj.bigSNP$map[-(2:3)]
names(map) <- c("chr", "pos", "a0", "a1")
info_snp <- snp_match(sumstats, map)
```

If no or few variants are actually flipped, you might want to disable the strand flipping option. Here, these are simulated data so all variants use the same strand and the same reference.

```{r}
info_snp <- snp_match(sumstats, map, strand_flip = FALSE)
df_beta <- info_snp[c("beta", "beta_se", "n_eff")]
```

## Computing LDpred2 scores for one chromosome

### Correlation

First, you need to compute correlations between variants.
We recommend to use a window size of 3 cM (see ref).

```{r}
POS2 <- snp_asGeneticPos(CHR, POS, dir = "tmp-data", ncores = 3)
```

```{r, cache=TRUE}
ind.chr <- which(info_snp$chr == 2)         ## indices in info_snp
ind.chr2 <- info_snp$`_NUM_ID_`[ind.chr]    ## indices in G
corr <- snp_cor(G, ind.col = ind.chr2, ncores = NCORES, 
                infos.pos = POS2[ind.chr2], size = 3 / 1000)
```

### Infinitesimal model

```{r, cache=TRUE}
(ldsc <- snp_ldsc2(corr, df_beta[ind.chr, ]))
h2_est <- ldsc[["h2"]]
```


```{r}
beta_inf <- snp_ldpred2_inf(corr, df_beta[ind.chr, ], h2 = h2_est)
```


```{r}
pred_inf <- big_prodVec(G, beta_inf, ind.row = ind.test, ind.col = ind.chr2)
AUC(pred_inf, y[ind.test])
```

### Grid of models

In practice, we recommend to test multiple values for h2 and p. 

```{r}
(h2_seq <- round(h2_est * c(0.7, 1, 1.4), 4))
(p_seq <- signif(seq_log(1e-4, 1, length.out = 17), 2))
(params <- expand.grid(p = p_seq, h2 = h2_seq, sparse = c(FALSE, TRUE)))
```

```{r, cache=TRUE, include=FALSE}
beta_grid <- snp_ldpred2_grid(corr, df_beta[ind.chr, ], params, ncores = NCORES,
                              burn_in = 20, num_iter = 50)
```
```{r, eval=FALSE}
# takes several minutes if you do not have many cores
beta_grid <- snp_ldpred2_grid(corr, df_beta[ind.chr, ], params, ncores = NCORES)
```


```{r}
pred_grid <- big_prodMat(G, beta_grid, ind.col = ind.chr2)
params$auc <- apply(pred_grid[ind.val, ], 2, AUC, target = y[ind.val])
```

```{r, out.width="90%", fig.asp=0.3}
library(ggplot2)
ggplot(subset(params, auc > 0.5), 
       aes(x = p, y = auc, color = as.factor(h2))) +
  theme_bigstatsr() +
  geom_point() +
  geom_line() +
  scale_x_log10(breaks = 10^(-5:0), minor_breaks = params$p) +
  facet_wrap(~ sparse, labeller = label_both) +
  labs(y = "AUC", color = "h2") +
  theme(legend.position = "top", panel.spacing = unit(1, "lines"))
```

```{r, message=FALSE}
library(dplyr)
params %>%
  mutate(sparsity = colMeans(beta_grid == 0), id = row_number()) %>%
  arrange(desc(auc)) %>%
  mutate_at(4:5, round, digits = 3) %>%
  slice(1:10)
```

You can then choose the best one according to your preferred criterion (e.g. max AUC) and evaluate it in the test set.

```{r}
best_grid_nosp <- params %>%
  mutate(id = row_number()) %>%
  filter(!sparse) %>% 
  arrange(desc(auc)) %>%
  slice(1) %>%
  pull(id) %>%
  beta_grid[, .]

AUC(big_prodVec(G, best_grid_nosp, ind.row = ind.test, ind.col = ind.chr2),
    y[ind.test])
```


```{r}
best_grid_sp <- params %>%
  mutate(id = row_number()) %>%
  filter(sparse) %>% 
  arrange(desc(auc)) %>%
  slice(1) %>%
  pull(id) %>%
  beta_grid[, .]

AUC(big_prodVec(G, best_grid_sp, ind.row = ind.test, ind.col = ind.chr2), 
    y[ind.test])
```

### Automatic model

```{r, cache=TRUE}
# takes a few minutes
auto <- snp_ldpred2_auto(corr, df_beta[ind.chr, ], h2_init = h2_est)
str(auto)
```

You should verify if the algorithm "converged".
This is not the case here, which is probably because the data is so small.

```{r}
plot_grid(
  qplot(y = auto$path_p_est) + 
    theme_bigstatsr() + 
    geom_hline(yintercept = auto$p_est, col = "blue") +
    scale_y_log10() +
    labs(y = "p"),
  qplot(y = auto$path_h2_est) + 
    theme_bigstatsr() + 
    geom_hline(yintercept = auto$h2_est, col = "blue") +
    labs(y = "h2"),
  ncol = 1, align = "hv"
)
```

```{r}
beta_auto <- auto$beta_est
pred_auto <- big_prodVec(G, beta_auto, ind.row = ind.test, ind.col = ind.chr2)
AUC(pred_auto, y[ind.test])
```

## Conclusion

We have seen how to run 3 versions of LDpred2 ("-inf", "-grid" and "-auto") for one chromosome.
You should do this for each chromosome and combine results.

### Reference

Coming soon on bioRxiv.

