---
title: "Computing polygenic scores using LDpred2"
author: "Florian Privé"
date: "February 24, 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

<center><iframe width="560" height="315" src="https://www.youtube.com/embed/aya8WsNAu6U" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe></center>

    

Here we show how to compute polygenic risk scores using [LDpred2](https://doi.org/10.1101/2020.04.28.066720).

**New:** if you install {bigsnpr} >= v1.10.4, LDpred2-grid and LDpred2-auto should be much faster for large data. You can install {bigsnpr} from CRAN using `install.packages("bigsnpr")`, or the latest version from GitHub using `remotes::install_github("privefl/bigsnpr")`.

This tutorial uses fake data for educational purposes only.
Another tutorial using another dataset can be found at https://privefl.github.io/bigsnpr-extdoc/polygenic-scores-pgs.html.

You should also probably look at [the code of the paper](https://github.com/privefl/paper-ldpred2/tree/master/code), particularly at [the code to prepare summary statistics (including performing the quality control presented in the Methods section "Quality control of summary statistics" of the paper)](https://github.com/privefl/paper-ldpred2/blob/master/code/prepare-sumstats.R), at [the code to read BGEN files into the data format used by bigsnpr](https://github.com/privefl/paper-ldpred2/blob/master/code/prepare-genotypes.R#L1-L62), and at [the code to prepare LD matrices](https://github.com/privefl/paper-ldpred2/blob/master/code/prepare-corr-spmat.R#L1-L26).

In practice, until we find a better set of variants, we recommend using the HapMap3 variants used in the PRS-CS and LDpred2 papers.
If you do not have enough data to use as LD reference (e.g. at least 2000 individuals), we provide an LD reference to be used directly ~~at https://doi.org/10.6084/m9.figshare.13034123~~, along with [an example R script](https://github.com/privefl/paper-ldpred2/blob/master/code/example-with-provided-ldref.R) on how to use it.
**New:** we now provide a new version of these LD references at https://doi.org/10.6084/m9.figshare.19213299 by forming independent LD blocks in the matrices, which can be useful for robustness and extra speed gains (see [this new paper](https://doi.org/10.1016/j.xhgg.2022.100136)).

Information about these variants can be retrieved with

```{r}
# $pos is in build GRCh37 / hg19, but we provide positions in 3 other builds 
info <- readRDS(runonce::download_file(
  "https://ndownloader.figshare.com/files/25503788",
  dir = "tmp-data", fname = "map_hm3_ldpred2.rds"))
str(info)
```

Note that you should **run LDpred2 genome-wide**; just build the SFBM (the sparse LD matrix on disk) so that it contains all 1M HapMap3 variants genome-wide (see the for-loop below).


## Downloading genotype data and summary statistics

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, width = 75, max.print = 30)
knitr::opts_knit$set(global.par = TRUE, root.dir = "..")
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dev = 'png')
```

You can download [the tutorial data](https://github.com/privefl/bigsnpr/raw/master/data-raw/public-data3.zip) and unzip files in R. We store those files in a directory called `"tmp-data"` here.

```{r}
# install.packages("runonce")
zip <- runonce::download_file(
  "https://github.com/privefl/bigsnpr/raw/master/data-raw/public-data3.zip",
  dir = "tmp-data")
unzip(zip)
```

First, you need to read genotype data from the PLINK files (or BGEN files) as well as the text file containing summary statistics.

```{r, echo=FALSE}
unlink(paste0("tmp-data/public-data3", c(".bk", ".rds")))
```

```{r}
# Load packages bigsnpr and bigstatsr
library(bigsnpr)
# Read from bed/bim/fam, it generates .bk and .rds files.
snp_readBed("tmp-data/public-data3.bed")
# Attach the "bigSNP" object in R session
obj.bigSNP <- snp_attach("tmp-data/public-data3.rds")
# See how the file looks like
str(obj.bigSNP, max.level = 2, strict.width = "cut")
# Get aliases for useful slots
G   <- obj.bigSNP$genotypes
CHR <- obj.bigSNP$map$chromosome
POS <- obj.bigSNP$map$physical.pos
y   <- obj.bigSNP$fam$affection
(NCORES <- nb_cores())
```

```{r}
# Read external summary statistics
sumstats <- bigreadr::fread2("tmp-data/public-data3-sumstats.txt")
str(sumstats)
```

We split the genotype data using part of the data to choose hyper-parameters and another part of the data to evaluate statistical properties of polygenic risk score such as AUC. Here we consider that there are 350 individuals to be used as validation set to tune hyper-parameters for LDpred2-grid. The other 153 individuals are used as test set to evaluate the final models.

```{r}
set.seed(1)
ind.val <- sample(nrow(G), 350)
ind.test <- setdiff(rows_along(G), ind.val)
```


## Matching variants between genotype data and summary statistics 

To match variants contained in genotype data and summary statistics, the variables `"chr"` (chromosome number), `"pos"` (physical genetic position in bp), `"a0"` (reference allele) and `"a1"` (alternative allele) should be available in the summary statistics and in the genotype data. These 4 variables are used to match variants between the two data frames. 
From the summary statistics, you need to get `"beta"`, `"beta_se"` (standard errors), and `"n_eff"` (the effective sample sizes per variant for a GWAS using logistic regression, and simply the sample size for continuous traits).

```{r, error=TRUE}
# sumstats$n_eff <- 4 / (1 / sumstats$n_case + 1 / sumstats$n_control)
# sumstats$n_case <- sumstats$n_control <- NULL
sumstats$n_eff <- sumstats$N
map <- setNames(obj.bigSNP$map[-3], c("chr", "rsid", "pos", "a1", "a0"))
df_beta <- snp_match(sumstats, map)
```

Here, there is problem with the matching; this is due to having different genome builds. You can either convert between builds with `snp_modifyBuild()` (or directly use the converted positions in `info`), or match by rsIDs instead.

```{r}
df_beta <- snp_match(sumstats, map, join_by_pos = FALSE)  # use rsid instead of pos
```

If no or few variants are actually flipped, you might want to disable the strand flipping option (`strand_flip = FALSE`) and maybe remove the few that were flipped (errors?).


## Computing LDpred2 scores genome-wide

**Some quality control on summary statistics is highly recommended (see paper and other tutorial).**
A new refined QC is described in [this new paper](https://doi.org/10.1016/j.xhgg.2022.100136).
See e.g. [the code](https://github.com/privefl/paper-misspec/tree/main/code) that was used to prepare the sumstats there.

### Correlation

First, you need to compute correlations between variants.
We recommend to use a window size of 3 cM (see [the LDpred2 paper](https://doi.org/10.1093/bioinformatics/btaa1029)).

```{r}
# POS2 <- snp_asGeneticPos(CHR, POS, dir = "tmp-data", ncores = NCORES)
# To avoid downloading "large" files, `POS2` has been precomputed
POS2 <- obj.bigSNP$map$genetic.dist
```

We create the on-disk sparse genome-wide correlation matrix on-the-fly:

```{r}
tmp <- tempfile(tmpdir = "tmp-data")

for (chr in 1:22) {
  
  # print(chr)
  
  ## indices in 'df_beta'
  ind.chr <- which(df_beta$chr == chr)
  ## indices in 'G'
  ind.chr2 <- df_beta$`_NUM_ID_`[ind.chr]
  
  corr0 <- snp_cor(G, ind.col = ind.chr2, size = 3 / 1000,
                   infos.pos = POS2[ind.chr2], ncores = NCORES)
  
  if (chr == 1) {
    ld <- Matrix::colSums(corr0^2)
    corr <- as_SFBM(corr0, tmp, compact = TRUE)
  } else {
    ld <- c(ld, Matrix::colSums(corr0^2))
    corr$add_columns(corr0, nrow(corr))
  }
}
```

To use the "compact" format for SFBMs, you need `packageVersion("bigsparser") >= package_version("0.5")`. 
Make sure to reinstall {bigsnpr} after updating {bigsparser} to this new version (to avoid crashes).

```{r}
file.size(corr$sbk) / 1024^3  # file size in GB
```

Note that you will need at least the same memory as this file size (to keep it cached for faster processing) + some other memory for all the results returned. If you do not have enough memory, processing will be very slow (because you would read the data from disk all the time). If using the one million HapMap3 variants, having 60 GB of memory should be enough.


### LDpred2-inf: infinitesimal model

```{r}
# Estimate of h2 from LD Score regression
(ldsc <- with(df_beta, snp_ldsc(ld, length(ld), chi2 = (beta / beta_se)^2,
                                sample_size = n_eff, blocks = NULL)))
h2_est <- ldsc[["h2"]]
```

```{r}
beta_inf <- snp_ldpred2_inf(corr, df_beta, h2 = h2_est)
```

```{r}
pred_inf <- big_prodVec(G, beta_inf, ind.row = ind.test, ind.col = df_beta[["_NUM_ID_"]])
pcor(pred_inf, y[ind.test], NULL)
```

LDpred2-inf would very likely perform worse than the other models presented hereinafter.
We actually recommend not to use it anymore.


### LDpred2(-grid): grid of models

In practice, we recommend to test multiple values for h2 and p. 

```{r}
(h2_seq <- round(h2_est * c(0.3, 0.7, 1, 1.4), 4))
(p_seq <- signif(seq_log(1e-5, 1, length.out = 21), 2))
(params <- expand.grid(p = p_seq, h2 = h2_seq, sparse = c(FALSE, TRUE)))
```

```{r}
set.seed(1)  # to get the same result every time
# takes less than 2 min with 4 cores
beta_grid <- snp_ldpred2_grid(corr, df_beta, params, ncores = NCORES)
```

```{r}
pred_grid <- big_prodMat(G, beta_grid, ind.col = df_beta[["_NUM_ID_"]])
params$score <- apply(pred_grid[ind.val, ], 2, function(x) {
  if (all(is.na(x))) return(NA)
  summary(lm(y[ind.val] ~ x))$coef["x", 3]
  # summary(glm(y[ind.val] ~ x, family = "binomial"))$coef["x", 3]
})
```

Note that missing values represent models that diverged substantially.

```{r, out.width="90%", fig.asp=0.5}
library(ggplot2)
ggplot(params, aes(x = p, y = score, color = as.factor(h2))) +
  theme_bigstatsr() +
  geom_point() +
  geom_line() +
  scale_x_log10(breaks = 10^(-5:0), minor_breaks = params$p) +
  facet_wrap(~ sparse, labeller = label_both) +
  labs(y = "GLM Z-Score", color = "h2") +
  theme(legend.position = "top", panel.spacing = unit(1, "lines"))
```

```{r, message=FALSE, warning=FALSE}
library(dplyr)
params %>%
  mutate(sparsity = colMeans(beta_grid == 0), id = row_number()) %>%
  arrange(desc(score)) %>%
  mutate_at(c("score", "sparsity"), round, digits = 3) %>%
  slice(1:10)
```

You can then choose the best model according to your preferred criterion (e.g. max AUC). Here, we use the Z-Score from the (linear or logistic) regression of the phenotype by the PRS since we have found it more robust than using the correlation or the AUC. It also enables adjusting for covariates in this step.

Also note that we separate both sparse and non-sparse models to show that their predictive performance are similar (in the paper). In practice, if you do not really care about sparsity, you could choose the best LDpred2-grid model among all sparse and non-sparse models. If you do, choose the best sparse one (if it is close enough to the best one).

```{r}
best_beta_grid <- params %>%
  mutate(id = row_number()) %>%
  # filter(sparse) %>% 
  arrange(desc(score)) %>%
  slice(1) %>%
  print() %>% 
  pull(id) %>% 
  beta_grid[, .]
```

```{r}
pred <- big_prodVec(G, best_beta_grid, ind.row = ind.test,
                    ind.col = df_beta[["_NUM_ID_"]])
pcor(pred, y[ind.test], NULL)
```


### LDpred2-auto: automatic model

We recommend to run many chains in parallel with different initial values for `p`. In [this new paper](https://doi.org/10.1016/j.xhgg.2022.100136), we have introduced two new parameters in LDpred2-auto for improving its robustness, `allow_jump_sign` and `shrink_corr`, and recommend to use them.

```{r}
set.seed(1)  # to get the same result every time
# takes less than 2 min with 4 cores
multi_auto <- snp_ldpred2_auto(corr, df_beta, h2_init = h2_est,
                               vec_p_init = seq_log(1e-4, 0.2, length.out = 30),
                               allow_jump_sign = FALSE, shrink_corr = 0.95,
                               ncores = NCORES)
str(multi_auto, max.level = 1)
str(multi_auto[[1]], max.level = 1)
```

You can verify whether the chains "converged" by looking at the path of the chains:

```{r}
library(ggplot2)
auto <- multi_auto[[1]]  # first chain
plot_grid(
  qplot(y = auto$path_p_est) + 
    theme_bigstatsr() + 
    geom_hline(yintercept = auto$p_est, col = "blue") +
    scale_y_log10() +
    labs(y = "p"),
  qplot(y = auto$path_h2_est) + 
    theme_bigstatsr() + 
    geom_hline(yintercept = auto$h2_est, col = "blue") +
    labs(y = "h2"),
  ncol = 1, align = "hv"
)
```

In the LDpred2 paper, we proposed an automatic way of filtering bad chains by comparing the scale of the resulting predictions (see [this code](https://github.com/privefl/paper-ldpred2/blob/master/code/run-ldpred2-gwide.R#L108-L112)). We have tested a somewhat equivalent and simpler alternative since, which we recommend here:

```{r}
# `range` should be between 0 and 1
(range <- sapply(multi_auto, function(auto) diff(range(auto$corr_est))))
(keep <- (range > (0.95 * quantile(range, 0.95))))
```

To get the final effects / predictions (after filtering):

```{r}
beta_auto <- rowMeans(sapply(multi_auto[keep], function(auto) auto$beta_est))
pred_auto <- big_prodVec(G, beta_auto, ind.row = ind.test, ind.col = df_beta[["_NUM_ID_"]])
```

```{r}
pcor(pred_auto, y[ind.test], NULL)
```


### lassosum2: grid of models

lassosum2 is a re-implementation of [the lassosum model](https://doi.org/10.1002/gepi.22050) that now uses the exact same input parameters as LDpred2 (`corr` and `df_beta`). It should be fast to run. It can be run next to LDpred2 and the best model can be chosen using the validation set.
Note that parameter 's' from lassosum has been replaced by a new parameter 'delta' in lassosum2, in order to better reflect that the lassosum model also uses L2-regularization (therefore, elastic-net regularization).

```{r}
beta_lassosum2 <- snp_lassosum2(corr, df_beta, ncores = NCORES)
```

```{r}
(params2 <- attr(beta_lassosum2, "grid_param"))
pred_grid2 <- big_prodMat(G, beta_lassosum2, ind.col = df_beta[["_NUM_ID_"]])
params2$score <- apply(pred_grid2[ind.val, ], 2, function(x) {
  if (all(is.na(x))) return(NA)
  summary(lm(y[ind.val] ~ x))$coef["x", 3]
  # summary(glm(y[ind.val] ~ x, family = "binomial"))$coef["x", 3]
})
```

```{r}
library(ggplot2)
ggplot(params2, aes(x = lambda, y = score, color = as.factor(delta))) +
  theme_bigstatsr() +
  geom_point() +
  geom_line() +
  scale_x_log10(breaks = 10^(-5:0)) +
  labs(y = "GLM Z-Score", color = "delta") +
  theme(legend.position = "top") +
  guides(colour = guide_legend(nrow = 1))
```

```{r}
library(dplyr)
best_grid_lassosum2 <- params2 %>%
  mutate(id = row_number()) %>%
  arrange(desc(score)) %>%
  print() %>% 
  slice(1) %>%
  pull(id) %>% 
  beta_lassosum2[, .]
```

```{r}
# Choose the best among all LDpred2-grid and lassosum2 models
best_grid_overall <- 
  `if`(max(params2$score, na.rm = TRUE) > max(params$score, na.rm = TRUE),
       best_grid_lassosum2, best_beta_grid)
```

```{r}
# Some cleaning
rm(corr); gc(); file.remove(paste0(tmp, ".sbk"))
```

## Inference with LDpred2-auto

LDpred2-auto has been very recently extended and improved to allow for estimating $h^2$, $p$, and $\alpha$, a third parameter that controls how expected effect sizes relate to minor allele frequencies. The new model assumed by LDpred2-auto is
\begin{equation}
\beta_j = S_j \gamma_j \sim \left\{
\begin{array}{ll}
\mathcal N \big( 0,~\sigma_\beta^2 \cdot (S_j^2)^{(\alpha + 1)} \big) & \mbox{with probability $p$,} \\
0 & \mbox{otherwise,}
\end{array}
\right.
\end{equation}
where $p$ is the proportion of causal variants, $h^2$ the (SNP) heritability, $\boldsymbol{\gamma}$ the effect sizes on the allele scale, $\boldsymbol{S}$ the standard deviations of the genotypes, and $\boldsymbol{\beta}$ the effects of the scaled genotypes.

Parameters $h^2$, $p$, and $\alpha$ (and 95% CIs) can be estimated using:
```{r}
coef_shrink <- 0.95
multi_auto <- snp_ldpred2_auto(corr, df_beta, h2_init = h2_est,
                               vec_p_init = seq_log(1e-4, 0.2, length.out = 50),
                               burn_in = 500, num_iter = 500, report_step = 20,
                               allow_jump_sign = FALSE, shrink_corr = coef_shrink,
                               ncores = NCORES)
(range <- sapply(multi_auto, function(auto) diff(range(auto$corr_est))))
(keep <- which(range > (0.95 * quantile(range, 0.95, na.rm = TRUE)) & range < 3))
```

```{r}
all_h2 <- sapply(multi_auto[keep], function(auto) tail(auto$path_h2_est, 500))
quantile(all_h2, c(0.5, 0.025, 0.975))
all_p <- sapply(multi_auto[keep], function(auto) tail(auto$path_p_est, 500))
quantile(all_p, c(0.5, 0.025, 0.975))
all_alpha <- sapply(multi_auto[keep], function(auto) tail(auto$path_alpha_est, 500))
quantile(all_alpha, c(0.5, 0.025, 0.975))
```

Predictive performance $r^2$ can also be inferred from the Gibbs sampler:
```{r}
bsamp <- lapply(multi_auto[keep], function(auto) auto$sample_beta)
all_r2 <- do.call("cbind", lapply(seq_along(bsamp), function(ic) {
  b1 <- bsamp[[ic]]
  Rb1 <- apply(b1, 2, function(x)
    coef_shrink * bigsparser::sp_prodVec(corr, x) + (1 - coef_shrink) * x)
  b2 <- do.call("cbind", bsamp[-ic])
  b2Rb1 <- as.matrix(Matrix::crossprod(b2, Rb1))
}))
quantile(all_r2, c(0.5, 0.025, 0.975))
```

To compare to
```{r}
beta_auto <- rowMeans(sapply(multi_auto[keep], function(auto) auto$beta_est))
pred_auto <- big_prodVec(G, beta_auto, ind.col = df_beta[["_NUM_ID_"]])
pcor(pred_auto, y, NULL)^2
```

These are not exactly the same, which we attribute to the small number of variants used in this fake data.

We can also get per-variant probabilities of being causal (for fine-mapping purposes)
```{r}
postp <- rowMeans(sapply(multi_auto[keep], function(auto) auto$postp_est))
qplot(y = postp, alpha = I(0.2)) + theme_bigstatsr()
```

You can also infer local per-block or per-annotation $h^2$ following the code [here](https://github.com/privefl/paper-infer/blob/3dcd2949f54d7f1fd5c346d1bbb35db4b84a719e/code/run-ldpred2-ukbb.R#L158-L165) and [here](https://github.com/privefl/paper-infer/blob/3dcd2949f54d7f1fd5c346d1bbb35db4b84a719e/code/height-meta-1600k.R#L170-L216).


## References

- Privé, F., Arbel, J., & Vilhjálmsson, B. J. (2020). [LDpred2: better, faster, stronger](https://doi.org/10.1093/bioinformatics/btaa1029). *Bioinformatics*, 36(22-23), 5424-5431.

- Privé, F., Arbel, J., Aschard, H., & Vilhjálmsson, B. J. (2022). [Identifying and correcting for misspecifications in GWAS summary statistics and polygenic scores](https://doi.org/10.1016/j.xhgg.2022.100136). *Human Genetics and Genomics Advances*, 3(4), 100136.
