---
title: "SCT"
author: "Florian Privé"
date: "May 23, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

In this document, I explain and show how to use SCT (Stacked Clumping and Thresholding).

Please also refer to and cite the corresponding paper:


## Data

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE, width = 75)
knitr::opts_knit$set(global.par = TRUE, root.dir = "..")
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', dev = 'png')
```

Download [data](https://github.com/privefl/bigsnpr/raw/master/data-raw/public-data.zip) and unzip files. I store those files in a directory called `"tmp-data"` here.

```{r, echo=FALSE, eval=FALSE}
unzip("data-raw/public-data.zip")
```

```{r, echo=FALSE}
unlink(paste0("tmp-data/public-data", c(".bk", ".rds")))
unlink(paste0("tmp-data/public-data-scores", c(".bk", ".rds")))
```

You can see [there](https://github.com/privefl/bigsnpr/blob/master/data-raw/public-data.R) how I generated these data from [the 1000 Genomes project](https://www.nature.com/articles/nature15393).

First, you need to read from the PLINK files (or BGEN files).

```{r}
# Load packages bigsnpr and bigstatsr
library(bigsnpr)
# Read from bed/bim/fam, it will create new files.
snp_readBed("tmp-data/public-data.bed")
# Attach the "bigSNP" object in R session
obj.bigSNP <- snp_attach("tmp-data/public-data.rds")
# See how it looks like
str(obj.bigSNP, max.level = 2, strict.width = "cut")
# Get aliases for useful slots
G   <- obj.bigSNP$genotypes
CHR <- obj.bigSNP$map$chromosome
POS <- obj.bigSNP$map$physical.pos
y   <- obj.bigSNP$fam$affection - 1
NCORES <- nb_cores()
# Check some counts for the 10 first SNPs
big_counts(G, ind.col = 1:10)
# Read external summary statistics
sumstats <- bigreadr::fread2("tmp-data/public-data-sumstats.txt")
str(sumstats)
```

```{r}
set.seed(1)
ind.train <- sample(nrow(G), 400)
ind.test <- setdiff(rows_along(G), ind.train)
```


## Steps in SCT

There are four steps in SCT that you can perform using functions `snp_match()` and `snp_grid_*()`.

### Matching variants

```{r, error=TRUE}
snp_match(sumstats, obj.bigSNP$map)
```

You need to have variables `"chr"`, `"pos"`, `"a0"` and `"a1"` in both summary statistics and variant information from the genotype data, as they are used to match variants between these two data frames.

```{r}
names(sumstats) <- c("chr", "rsid", "pos", "a0", "a1", "beta", "p")
map <- obj.bigSNP$map[-(2:3)]
names(map) <- c("chr", "pos", "a0", "a1")
info_snp <- snp_match(sumstats, map)
```

If no or few variants are actually flipped, you might want to disable the strand flipping option.

```{r}
info_snp <- snp_match(sumstats, map, strand_flip = FALSE)
beta <- info_snp$beta
lpval <- -log10(info_snp$p)
```

### Clumping

First, you need to compute sets of clumping corresponding to many hyper-parameters (threshold of correlation for clumping, window size of LD checking, and possibly imputation accuracy threshold).

```{r, cache=TRUE}
all_keep <- snp_grid_clumping(G, CHR, POS, ind.row = ind.train,
                              lpS = lpval, ncores = NCORES)
str(all_keep)  ## only 3 chromosomes in these data
```

Here, we use 28 (7 x 4) different sets of hyper-parameters for clumping.

```{r}

attr(all_keep, "grid")
```

### Thresholding

Then, for each chromosome, for each set of clumping indices previously computed and for each p-value threshold, the corresponding C+T score is computed.

```{r}
multi_PRS <- snp_grid_PRS(G, all_keep, beta, lpval, ind.row = ind.train,
                          backingfile = "tmp-data/public-data-scores", 
                          n_thr_lpS = 50, ncores = NCORES)
dim(multi_PRS)  ## 4200 C+T scores
```

### Stacking

A penalized regression is finally used to learn a linear combination of all C+T scores. 

```{r}
final_mod <- snp_grid_stacking(multi_PRS, y[ind.train], ncores = NCORES, K = 4)
summary(final_mod$mod)
plot(final_mod$mod)
```

For options of fitting penalized regressions, see [this vignette](https://privefl.github.io/bigstatsr/articles/penalized-regressions.html).

From stacking C+T scores, we can derive a unique vector of weights.

```{r}
new_beta <- final_mod$beta.G
ind <- which(new_beta != 0)
```

```{r}
library(ggplot2)
ggplot(data.frame(y = new_beta, x = beta)[ind, ]) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  geom_abline(slope = 0, intercept = 0, color = "blue") +
  geom_point(aes(x, y), size = 0.6) +
  theme_bigstatsr() +
  labs(x = "Effect sizes from GWAS", y = "Non-zero effect sizes from SCT")
```

We can use this vector of SNP weights to predict in another set.

```{r}
pred <- final_mod$intercept + 
  big_prodVec(G, new_beta[ind], ind.row = ind.test, ind.col = ind)
ggplot(data.frame(
  Phenotype = factor(y[ind.test], levels = 0:1, labels = c("Control", "Case")),
  Probability = 1 / (1 + exp(-pred)))) + 
  theme_bigstatsr() +
  geom_density(aes(Probability, fill = Phenotype), alpha = 0.3)
AUCBoot(pred, y[ind.test])
```

## Best C+T score

If you want to compute the best C+T score based on the computed grid, you can do:

```{r, message=FALSE}
library(tidyverse)
grid2 <- attr(all_keep, "grid") %>%
  mutate(thr.lp = list(attr(multi_PRS, "grid.lpS.thr")), num = row_number()) %>%
  unnest()
s <- nrow(grid2)
grid2$auc <- big_apply(multi_PRS, a.FUN = function(X, ind, s, y.train) {
  # Sum over all chromosomes, for the same C+T parameters
  single_PRS <- rowSums(X[, ind + s * (0:2)])  ## replace by 0:21 in real data
  bigstatsr::AUC(single_PRS, y.train)
}, ind = 1:s, s = s, y.train = y[ind.train],
a.combine = 'c', block.size = 1, ncores = NCORES)
```

```{r}
max_prs <- grid2 %>% arrange(desc(auc)) %>% slice(1:10) %>% print() %>% slice(1)
```

```{r, fig.asp=2.5, out.width="50%"}
ggplot(grid2) +
  geom_point(aes(thr.lp, auc)) +
  # facet_grid(thr.imp ~ thr.r2 + size) +
  facet_grid(thr.r2 + size ~ .) +
  scale_x_log10(limits = c(1, NA)) +
  ylim(0.55, NA) +
  theme_bigstatsr(size.rel = 0.6) +
  labs(x = "-log10(p-value) threshold (log scale)", y = "AUC")
```

```{r}
ind.keep <- unlist(map(all_keep, max_prs$num))
sum(lpval[ind.keep] > max_prs$thr.lp)
AUCBoot(
  snp_PRS(G, beta[ind.keep], ind.test = ind.test, ind.keep = ind.keep,
          lpS.keep = lpval[ind.keep], thr.list = max_prs$thr.lp),
  y[ind.test]
)
```

