---
title: "Principal Component Analysis"
author: "Florian Privé"
date: "November 13, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

```{r}
library(bigsnpr)
library(ggplot2)
```

## Data

Let us use a subsetted version of the 1000 Genomes project data we provide. Some quality control has already been done; otherwise, you can use `snp_plinkQC()`.

```{r}
bedfile <- download_1000G("data")
```

## Relatedness

First, let us detect all pairs of related individuals.

```{r}
plink2 <- download_plink2("data")
rel <- snp_plinkKINGQC(
  plink2.path = plink2,
  bedfile.in = bedfile,
  thr.king = 2^-4.5,
  make.bed = FALSE,
  ncores = nb_cores()
)
str(rel)
```

## Principal Component Analysis (PCA)

We then compute PCA without using the related individuals. Function `bed_autoSVD()` should take care of Linkage Disequilibrium (LD).

```{r}
(obj.bed <- bed(bedfile))
ind.rel <- match(c(rel$ID1, rel$ID2), obj.bed$fam$sample.ID)
ind.norel <- rows_along(obj.bed)[-ind.rel]

obj.svd <- bed_autoSVD(obj.bed, ind.row = ind.norel, k = 20, 
                       ncores = nb_cores())
```

## Outlier sample detection

Then, we look at if there are individual outliers.

```{r}
prob <- bigutilsr::prob_dist(obj.svd$u, ncores = nb_cores())
S <- prob$dist.self / sqrt(prob$dist.nn)

ggplot() +
  geom_histogram(aes(S), color = "#000000", fill = "#000000", alpha = 0.5) +
  scale_x_continuous(breaks = 0:5 / 5, limits = c(0, NA)) +
  scale_y_sqrt(breaks = c(10, 100, 500)) +
  theme_bigstatsr() +
  labs(x = "Statistic of outlierness", y = "Frequency (sqrt-scale)")

plot_grid(plotlist = lapply(7:10, function(k) {
  plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
    aes(color = S) +
    scale_colour_viridis_c()
}), scale = 0.95)

plot_grid(plotlist = lapply(7:10, function(k) {
  plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
    aes(color = S > 0.5) +  # threshold based on histogram
    scale_colour_viridis_d()
}), scale = 0.95)
```

## PCA without outlier

We recompute PCA without outliers, starting with the previous set of variants kept (we can therefore skip the initial clumping step).

```{r}
ind.row <- ind.norel[S < 0.5]
ind.col <- attr(obj.svd, "subset")
obj.svd2 <- bed_autoSVD(obj.bed, ind.row = ind.row, 
                        ind.col = ind.col, thr.r2 = NA,
                        k = 20, ncores = nb_cores())
```

## Verification

```{r}
plot(obj.svd2)
```


```{r, out.width="95%"}
plot(obj.svd2, type = "loadings", loadings = 1:20, coeff = 0.4)
```


```{r, out.width="80%"}
plot(obj.svd2, type = "scores", scores = 1:20, coeff = 0.4)
```

## Project remaining individuals

```{r}
PCs <- matrix(NA, nrow(obj.bed), ncol(obj.svd2$u))
PCs[ind.row, ] <- predict(obj.svd2)

proj <- bed_projectSelfPCA(obj.svd2, obj.bed, 
                           ind.row = rows_along(obj.bed)[-ind.row],
                           ncores = 1) # useless -> too few individuals
PCs[-ind.row, ] <- proj$OADP_proj
```

```{r}
plot(PCs[ind.row, 7:8], pch = 20, xlab = "PC7", ylab = "PC8")
points(PCs[-ind.row, 7:8], pch = 20, col = "blue")
```

## References
