---
title: "How to capture Population Structure from PCA?"
author: "Florian Privé"
date: "October 5, 2017"
output: html_document
bibliography: refs.bib
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align = "center", dev = "png", cache = TRUE,
                      fig.asp = 0.7, out.width = "70%")
```

In this vignette, I will show how to perform Principal Component Analysis (PCA) with packages **bigstatsr** and **bigsnpr** [@Prive2017]. I'll also show why pruning and removing long-range Linkage Disequilibrium (LD) are two important steps before computing PCs in order to capture Population Structure, as reported in [@Abdellaoui2013].

## Data

I'll use data from a case/control cohort on celiac disease [@Dubois2010]. The data has already been QCed, genotyped SNPs have been imputed and binary PLINK files have been converted to the "bigSNP" format used in **bigsnpr** (see these [preprocessing steps](https://www.biorxiv.org/content/biorxiv/suppl/2017/09/19/190926.DC1/190926-1.html)).

```{r, cache=FALSE}
library(bigsnpr)
library(ggplot2)

celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- nb_cores()

# "Verification" there is no missing value
big_counts(G, ind.col = 1:12) # OK

# Get population from external files
pop.files <- list.files(path = "data", pattern = "cluster_*", full.names = TRUE)
pop <- snp_getSampleInfos(celiac, pop.files)[[1]]
pop.names <- c("Netherlands", "Italy", "UK1", "UK2", "Finland")
```

## Principal Component Analysis

### On the whole genotype matrix

```{r}
svd1 <- big_randomSVD(G, snp_scaleBinom(), ncores = NCORES)
```

```{r}
plot(svd1, type = "scores") +
  aes(color = pop.names[pop]) +
  labs(color = "Population")
```

First two PCs capture Population Structure.

```{r}
plot(svd1, type = "scores", scores = 3:4) +
  aes(color = pop.names[pop]) +
  labs(color = "Population")
```

PC3 and PC4 don't capture Population Structure.

```{r}
# The SNP with max loading for PC3
theone1.1 <- which.max(abs(svd1$v[, 3]))
# The SNP with max loading for PC4
theone1.2 <- which.max(abs(svd1$v[, 4]))
plot(svd1, type = "scores", scores = 3:4) +
  aes(color = as.factor(paste(G[, theone1.1], G[, theone1.2]))) +
  labs(color = "Genotypes")
```

So basically, PC3 is capturing a variation of one SNP, as well as PC4. These two SNPs are located in long-range LD regions and corresponds to peaks in the next figure.

```{r}
plot(svd1, type = "loadings", loadings = 1:4, coeff = 0.7)
```


### When pruning only

In fact, I'm using clumping on the Minor Allele Frequencies (MAF) instead of pruning. You can see [this vignette](https://privefl.github.io/bigsnpr/articles/pruning-vs-clumping.html) to know why.

```{r}
ind.keep2 <- snp_clumping(G, CHR, ncores = NCORES)
svd2 <- big_randomSVD(G, snp_scaleBinom(), ncores = NCORES,
                      ind.col = ind.keep2)
```

```{r}
plot(svd2, type = "scores", scores = 3:4) +
  aes(color = pop.names[pop]) +
  labs(color = "Population")
```

```{r}
theone2 <- ind.keep2[which.max(abs(svd2$v[, 4]))]
plot(svd2, type = "scores", scores = 3:4) +
  aes(color = as.factor(G[, theone2])) +
  labs(color = "Genotype")
```

PC4 captures LD.

### When removing long-range LD regions only

As recommended in [@Price2008a].

```{r}
ind.keep3 <- cols_along(G)[-snp_indLRLDR(CHR, POS)]
svd3 <- big_randomSVD(G, snp_scaleBinom(), ncores = NCORES,
                      ind.col = ind.keep3)
```

```{r}
plot(svd3, type = "scores", scores = 3:4) +
  aes(color = pop.names[pop]) +
  labs(color = "Population")
```

This is quite better at capturing Population Structure. Yet..

```{r, fig.asp=1.2}
plot(svd3, type = "loadings", loadings = 1:10, coeff = 0.6)
```

So, first 4 PCs are mostly capturing Population Structure, but the next PCs are likely to capture only LD structure.

```{r}
plot(svd3, type = "scores", scores = 5:6) +
  aes(color = pop.names[pop]) +
  labs(color = "Population")
```

```{r}
theone3 <- ind.keep3[which.max(abs(svd3$v[, 5]))]
plot(svd3, type = "scores", scores = 5:6) +
  aes(color = as.factor(G[, theone3])) +
  labs(color = "Genotype")
```

### When pruning and removing long-range LD regions

As recommend in [@Abdellaoui2013].

```{r}
ind.keep4 <- snp_clumping(G, CHR, ncores = NCORES,
                          exclude = snp_indLRLDR(CHR, POS))
svd4 <- big_randomSVD(G, snp_scaleBinom(), ncores = NCORES,
                      ind.col = ind.keep4)
```

```{r}
plot(svd4, type = "scores", scores = 3:4) +
  aes(color = pop.names[pop]) +
  labs(color = "Population")
```

```{r}
plot(svd4, type = "scores", scores = 5:6) +
  aes(color = pop.names[pop]) +
  labs(color = "Population")
```

Maybe, PC5 is now capturing some non-European ancestry.

```{r, fig.asp=1.2}
plot(svd4, type = "loadings", loadings = 1:10, coeff = 0.6)
```

The loadings are all approximately normally distributed (no huge peak), we're good.

### An automatic procedure

You can use either the previous method (for human data only) or try the following automatic procedure. You'll find details in [@Prive2017].

```{r}
svd0 <- snp_autoSVD(G, CHR, POS, ncores = NCORES)
attr(svd0, "lrldr")
```

```{r}
plot(svd0, type = "scores", scores = 3:4) +
  aes(color = pop.names[pop]) +
  labs(color = "Population")
```

```{r}
plot(svd0, type = "scores", scores = 5:6) +
  aes(color = pop.names[pop]) +
  labs(color = "Population")
```

```{r, fig.asp=1.2}
plot(svd0, type = "loadings", loadings = 1:10, coeff = 0.6)
```

## Conclusion

Always use both pruning and removing of long-range LD regions when computing the PCs, as recommended in [@Abdellaoui2013]. To check that the results are capturing Population Structure, you should plot PCA scores. To check that PCs are not capturing LD, you should check PCA loadings.

## References
