---
title: "How to capture Population Structure from PCA?"
author: "Florian Privé"
date: "October 5, 2017"
output: html_document
bibliography: ../../articles/library.bib
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align = "center", dev = "png", cache = TRUE,
                      fig.asp = 0.7, out.width = "70%")
```

In this vignette, I will show how to perform Principal Component Analysis (PCA) with packages **bigstatsr** and **bigsnpr** [@Prive2017] and why pruning and removing Long-range LD are two important steps of computing PCs for capturing Population Structure (as reported in [@Abdellaoui2013]).

## Data

I'll use data from a case/control cohort on celiac disease [@Dubois2010]. The data has already been QCed, genotyped SNPs have been imputed and binary PLINK files have been converted to the "bigSNP" format used in **bigsnpr** (see the [preprocessing steps](https://www.biorxiv.org/content/biorxiv/suppl/2017/09/19/190926.DC1/190926-1.html).

```{r}
library(bigsnpr)

celiac <- snp_attach("backingfiles/celiacQC.rds")
G <- celiac$genotypes
CHR <- celiac$map$chromosome
POS <- celiac$map$physical.pos
NCORES <- nb_cores()

# "Verification" there is no missing value
big_counts(G, ind.col = 1:15) # OK

# Get population from external files
pop.files <- list.files(path = "data", pattern = "cluster_*", full.names = TRUE)
pop <- snp_getSampleInfos(celiac, pop.files)[[1]]
pop.names <- c("Netherlands", "Italy", "UK1", "UK2", "Finland")
```



## References
