---
title: "LD scores and 'loadings'"
author: "Florian Privé"
date: "January 15, 2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
```

```{r}
library(bigsnpr)
library(ggplot2)
```

## Data

Let us use a subsetted version of the 1000 Genomes project data we provide. Some quality control has already been done; otherwise, you can use `snp_plinkQC()`.

```{r}
bedfile <- download_1000G("tmp-data")
```

## Relatedness

First, let us detect all pairs of related individuals.

```{r}
plink2 <- download_plink2("tmp-data")
rel <- snp_plinkKINGQC(
  plink2.path = plink2,
  bedfile.in = bedfile,
  thr.king = 2^-4.5,
  make.bed = FALSE,
  ncores = nb_cores()
)
str(rel)
```

## Principal Component Analysis (PCA)

We then compute PCA without using the related individuals. Function `bed_autoSVD()` should take care of Linkage Disequilibrium (LD). We have to compute the PCA while removing all variants from one chromosome. We can still precompute the pruning step to do it only once.

```{r}
(obj.bed <- bed(bedfile))
fam2 <- bigreadr::fread2(sub_bed(bedfile, ".fam2"))
ind.row <- which(!obj.bed$fam$sample.ID %in% rel$IID2 &
                   fam2$`Super Population` == "EUR")

ind.keep <- bed_clumping(obj.bed, ind.row = ind.row, ncores = nb_cores())
```

## Computing LD scores and "loadings"

```{r}
CHR <- obj.bed$map$chromosome
POS <- obj.bed$map$physical.pos

library(bigparallelr)
registerDoParallel(cl <- makeCluster(nb_cores()))
all_stats <- foreach(chr = 1:22, .combine = "rbind") %dopar% {
  ## LD scores
  ind.chr <- which(CHR == chr)
  library(bigsnpr)
  ## Should implement bed_cor() directly
  G <- snp_attach(snp_readBed2(bedfile, tempfile(), ind.row, ind.chr))$genotypes
  LD <- snp_cor(G, infos.pos = POS[ind.chr])  # more lenient alpha?
  library(Matrix)
  LD_scores <- colSums(LD ** 2)
  ## PCA
  obj.svd <- bed_autoSVD(
    obj.bed, ind.row = ind.row, k = 10, thr.r2 = NA,
    ind.col = ind.keep[CHR[ind.keep] != chr])
  ## Z-scores (association with PC scores) ~ loadings
  Z <- bigsnpr:::multLinReg(obj.bed, ind.row, ind.chr, U = obj.svd$u)
  cbind(LD_scores, Z)
}
stopCluster(cl)
saveRDS(all_stats, "tmp-data/ldscores-loadings.rds")
```

```{r}
LD_scores <- all_stats[, 1]
Z2 <- all_stats[, -1] ** 2
cor(Z2, LD_scores, use = "pairwise.complete.obs")
 #  0.066223166
 #  0.058258742
 # -0.002232321
 #  0.024427324
 # -0.029319222
 # -0.007915754
 # -0.012346952
 # -0.018130519
 # -0.010315609
 #  0.001048635

bin_plot <- function(x, N = 10) {
  grp <- cut(x, quantile(x, seq_len(N) / N, na.rm = TRUE))
  plot(by(LD_scores, grp, median), by(x, grp, mean))
}
bin_plot(Z2[, 1])
bin_plot(Z2[, 2])
bin_plot(Z2[, 3])
bin_plot(Z2[, 4])
bin_plot(Z2[, 5])
bin_plot(Z2[, 6])
bin_plot(Z2[, 7])
bin_plot(Z2[, 8])
```

